---
title: "R Tutorial"
subtitle: "An introduction to R"
date: "2020-09-01"
author: "Derek Michael Wright <wrightmderek@gmail.com> [www.dblogr.com/](https://dblogr.netlify.com/)"
math: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

---

# Download R

Download R here
Download RStudio here

# Introduction

R can be used as a caclulator

```{r}
5 + 3 # = 8
24 / (1 + 2) # = 24 / 3 = 8
2^3 # = 2 * 2 * 2 = 8
sqrt(64) # 8 * 8 = 64
-log10(0.05 / 5000000) # = 8
```

Information can be stored in user defined objects. e.g., xx

```{r}
xx <- -log10(0.05 / 5000000)
xx
```

R has many useful built in functions

```{r}
1:10
as.character(1:10)
rep(1:2, times = 5)
rep(1:5, times = 2)
rep(1:5, each = 2)
rep(1:5, length.out = 7)
help(rep)
seq(5, 50, by = 5)
seq(5, 50, length.out = 5)
mean(1:10)
sd(1:10)
min(1:10)
max(1:10)
range(1:10)
```

Users can also create their own functions

```{r}
customFunction <- function(x) {
  mean(x) + sd(x)
}
customFunction(1:10)
```

Objects can be stored in multiple forms:

- c()
- matrix()
- data.frame()
- list()

Data can be saved in many forms:

- numeric
- integer
- character
- factor

```{r}
x1 <- 1:10
x1
x1 <- c(1,2,3,4,5,6,7,8,9,10)
x1
x2 <- matrix(1:100, nrow = 10, ncol = 10, byrow = T)
x2
x2 <- matrix(1:100, nrow = 10, ncol = 10, byrow = F)
x2
x3 <- data.frame(
  y1 = 1:10,
  y2 = rep(c("1","2"), times = 5),
  y3 = c(1,1,1,1,1,2,2,2,3,3),
  y4 = rep(1:5, times = 2),
  y5 = rep(1:5, each = 2),
  y6 = seq(5, 50, by = 5),
  y7 = log10(1:10),
  y8 = (1:10)^3
)
x3[1:5,1:5]
x3[,"y3"]
x3$y3
x3$y3 <- as.character(x3$y3)
x3$y3
x3$y3 <- as.numeric(x3$y3)
x3$y3
x3$y3 <- as.factor(x3$y3)
x3$y3
x3$y3 <- factor(x3$y3, levels = c("3","2","1"))
x3$y3
x3$y2 <- factor(x3$y2)
x3$y2
xx <- list(x1, x2, x3)
```

Internal structure of an object can be checked with str()

```{r}
str(x1)
str(x2)
str(x3)
str(xx)
```

Additional libraries can be installed and loaded for use.

```{r eval = F}
install.packages("scales")
```

```{r}
library(scales)
rescale(1:10, to = c(1,30))
```

# Plotting

```{r}
plot(x = x3$y1, y = x3$y5)
plot(x = x3$y1, y = x3$y5, col = x3$y2)
plot(x = x3$y7, y = x3$y8, type = "both")
library(ggplot2)
mp <- ggplot(x3, aes(x = y7, y = y8))
mp + geom_point()
mp + geom_point(aes(color = y2))
mp + geom_line()
mp + geom_line(aes(color = y2))
mp + geom_smooth()
mp + geom_smooth(method = "loess")
mp + geom_smooth(method = "lm")
mp <- ggplot(x3, aes(x = y2, y = y7))
mp + geom_boxplot()
mp + geom_boxplot() + geom_point()
mp <- ggplot(x3, aes(x = y3))
mp + geom_histogram(stat = "count")
```

Statistics

Under construction…